---
title: "VCFArrow"
author: "Tomas Hrbek"
date: "06/06/2026"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{VCFArrow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


# Installation

This package needs to be installed from GitHub.
`devtools::install_github("legalLab/VCFArrow")`

The package is build around Apache Arrow (https://arrow.apache.org/).
The Apache Arrow libarrow-dev, libparquet-dev, libarrow-dataset-dev, and libarrow-acero-dev libraries need to be installed.
Installation instruction are found here (https://arrow.apache.org/install/).


# Introduction

This package has six main functionalities: (1) reading in and writing out VCF data files, including precalculation of diverse metrics used in the filtering functions; (2) data filtering functions; (3) merging VCFs; (4) subsetting VCFs; (5) generation of summary statistics and graphs; and (6) data transformation and writing out of other data formats ( population genetic and phylogenetic formats) for downstream analyses.
This package is a reimagination of the `vcf2others` package. Same as `vcf2others`, it uses the Apache Arrow library functions to manipulate SNP data via dplyr verb equivalents. However, unlike `vcf2others`, `VCFArrow` now reads in the VCF data file into an S4 R object constructed around an Apache Arrow data structure. It precalculate a number of metrics and stores them in the S4 object. The VCF genotype matrix is stored as a long tidy table, and the data in the S4 object are lazy loaded. Many common tasks have been delegated to shared, C++ backed APIs and functions. These modifications and optimizations now permit extremely fast data filtering, manipulation and transformation. It is possible to process VCFs comprising tens of millions of SNVs and hundreds of individuals in matter minutes with minimal RAM overhead.

## Meta

* Please [report here any issues or bugs or suggestions](https://github.com/legalLab/VCFArrow/issues).
* License: MIT.
* Get citation information for `VCFArrow` in R by running `citation(package='VCFArrow')`.


# How to use the functions of this package

Following are examples of the usage of the functions of this package.

## Load VCF and associated files of individuals and groups

First we need to read in the VCF. If one has information on the grouping of individuals, this information can also be put into the VCFArrow object. This grouping information is stored in a 'strata' file (a tab-separated text file with two columns with an 'id' and 'pop' headers). The grouping of individuals can be used in filtering steps, and it is necessary for transformation of the VCFArrow object to other population genetic formats, many of which require it.

The data used in this example are from Mota et al. (2026)---a population genomic analysis of the *Phyllomedusa vaillantii* species complex. The original dataset has 18579683 SNVs by 56 samples. For the purpose of this tutorial, the dataset was randomly subsampled to 10000 SNVs by 18 samples (five samples by three main lineages of *Phyllomedusa vaillantii* as ingroups, plus three samples of *Phyllomedusa bicolor* included as outgroups).

```{r setup} 
library(VCFArrow)
library(dplyr)

# set path to example files and project name
data_path <- system.file("extdata", package="VCFArrow")
project <- "vaillantii_"
postfix <- "discosnp_sub"

# load vcf and assign individuals to groups based on 'strata'
vcf <- read_vcf(file.path(data_path, paste0(project, postfix, ".vcf.gz"))) |>
  set_vcf_groups(data_path)

# visualize the VCFArrow object
vcf

# read individuals to include
# if indivs is blank, the default is to use all individuals
indivs <- read.table(file.path(data_path, "indivs_b"), header = TRUE)$id
  
# check if all samples are in VCF sample names
if (any(!(indivs %in% vcf@samples))) stop(paste("Some individuals in list not in VCF"))

```


## Assessment of missing data

The function `assess_vcf_missing_data()` is used to generate a table and graph of missing data for each sample in a VCFArrow object, while the `assess_vcf_coverage()` generates a violin plot of read depths of each sample. Both are useful for visualizing data before and after filtering to evaluate the effect of filtering. The functions accept parameters which form part of the name of the output file. The idea is that the output file name contains information on the name of the project (project), how the VCF was extracted (postfix), and how it was filtered (fltr). The 'postfix' and 'fltr' can be left blank if the VCF name does not contain this information. Samples are ordered and colored by group assignment. The 'details' flag specifies whether or not filtering information is presented in the figure or if only the species name is reported. The default is to report species name and details.

```{r} 
# define results path
res_path <- data_path
# species for plot title
species <- "Phyllomedusa vaillantii"
# filter - filter parameters in file name
fltr <- ""

assess_vcf_missing_data(vcf, res_path, species, paste0(project, postfix, fltr))
assess_vcf_coverage(vcf, res_path, species, paste0(project, postfix, fltr))

```

## Basic stats for individuals in VCF

The function `vcf_stats()` is used to generate a table with basic stats for each sample in the VCFArrow object. These stats include per sample average read depth, heterozygosity, number of heterozygotes, homozygotes, REF homozygotes and ALT homogygotes, percent missing SNVs, total missing SNVs, total non-missing SNVs and total SNVs, Watterson's Theta and Pi for the entire dataset, and per group Watterson's Theta and Pi, plus the number of individuals in each group. The calculation of Watterson's Theta and Pi can be included or excluded based on the 'theta' flag. 

``` {r}
# get a table of basic sample stats, including Watterson's Theta and Pi

vcf_stats(vcf, res_path, paste0(project, postfix, fltr), theta = TRUE)
```


## Filtering, subsetting, merging and otherwise wrangling VCF files

Most but not all functions can be used by the user directly on the `VCFArrow` object. Some functions, such as `.vcf_filter_rows()` and `.vcf_filter_columns()` are common APIs used by other functions to perform filtering. Other functions, such as `vcf_filter_invariant()` will remove invariant SNPs from a VCFArrow object; however, VCF by definition should not have invariant SNPs. So this function is primarily called by other functions to remove loci that may have become invariant following filtering.

### Subsetting samples and variants

There are two types of subsetting functions. The first type is used to subset the VCFArrow object by specific samples  `vcf_extract_samples()` or groups (all samples of a group) of samples `vcf_extract_groups()`. The 'keep' flag in both functions determines whether the indicated samples should be extracted from the VCF ('keep = TRUE') or excluded from the VCF ('keep = FALSE'). Once subsetted, the invariant SNVs are removed by default, however, this behavior is controlled by the 'f_invar' flag. The second type subsets a random number of SNVs from the VCF. The subsetting is done either across the entire VCF `vcf_sub_SNVs()` or proportionately within each chromosome/linkage group `vcf_sub_SNVs_stratified()`. For repeatability, both functions accept a 'seed'; if 'seed' is not specified, it is randomly generated.

```{r} 
# subset VCF by individuals (default keep = TRUE, remove invariants f_invar = TRUE)
indivs1 <- c("Pv14", "Pv27", "Pv28", "Pv78", "Pv126")
vcf1 <- vcf_extract_samples(vcf, indivs1)

# subset VCF by groups (default keep = TRUE, remove invariants f_invar = TRUE)
groups1 <- c("GS", "BS", "WA")
vcf1 <- vcf_extract_groups(vcf, groups1)

# subset VCF by a random  number of variants (default n_SNVs = 10000)
vcf1 <- vcf_sub_SNVs(vcf, n_SNVs = 5000)

# subset VCF by a random  number of variants proportional to chromosome length (default n_SNVs = 10000)
# should be used when SNVs are mapped to chromosomes or scaffolds
vcf1 <- vcf_sub_SNVs_stratified(vcf, n_SNVs = 5000)

```

### Filtering

Filtering consists of removing samples and variants based on specific properties. The function `vcf_filter_missing()` is the only function that filters samples based on % missing SNV threshold. The other functions filter on properties of individuals SNVs across samples. The function `vcf_filter_missingness()` removes SNVs above a % missing threshold, `vcf_filter_quality()` removes SNVs below a quality threshold, `vcf_filter_pass()` removes SNVs that did not PASS filters, `vcf_filter_maf()` removes SNVs with minor allele frequency below a % threshold, `vcf_filter_hets()` removes SNVs with heterozigosity above a % threshold, `vcf_filter_rank()` removes SNPs with rank below a % threshold (rank is specific to VCFs generated by DiscoSnpRad and measures the likelihood of observed SNP variation being due to paralogs), `vcf_filter_adr()` either removes (make genotype missing) or corrects (make genotype homozygous) genotypes with allele ratio imbalance, `vcf_filter_coverage()` removes SNVs below a specific coverage (make genotypes missing), and `vcf_filter_biallelic()` removes all variants that are not biallelic. Finally the `vcf_filter_invariant()` removes invariant SNVs (SNVs may become invariant after subsetting and filtering). The function `vcf_filter_indels()` removes indels from the VCF, retaining only SNPs. The functions `vcf_filter_oneSNV()` and `vcf_filter_multiSNV()` filter the VCF to remove linked SNVs (retain only unliked SNPs) or retain only linked SNVs, respectively. In general, all analyses assume that SNVs are unlinked, however, some analyses such as FineRadStructure work with linked SNVs.


```{r} 
# filter VCF for analyses (unlinked SNPs)
vcf_oneSNP <- vcf_extract_samples(vcf, indivs) |>
  vcf_filter_indels() |>
  vcf_filter_biallelic() |>
  vcf_filter_maf(.03) |>
  vcf_filter_coverage(10) |>
  vcf_filter_missingness(.2) |>
  vcf_filter_missing(.3) |>
  vcf_filter_oneSNV()

# see how many variants remained
nrow(vcf_oneSNP@variants)

# filter VCF for analyses (linked SNPs)
vcf_multiSNP <- vcf_extract_samples(vcf, indivs) |>
  vcf_filter_indels() |>
  vcf_filter_biallelic() |>
  vcf_filter_maf(.03) |>
  vcf_filter_coverage(10) |>
  vcf_filter_missingness(.2) |>
  vcf_filter_missing(.3) |>
  vcf_filter_multiSNV()

# see how many variants remained
nrow(vcf_multiSNP@variants)

```

### Extracting and binding

The functions `vcf_merge_bind()`, `vcf_bind_sparse()`, `vcf_extract_samples()` and `vcf_extract_groups()` are used to merge two or more VCFArrow objects, and to extract samples/groups of samples from a VCFArrow objects storing them in a separate VCFArrow object. 

```{r} 
# extract individuals from VCFArrow object (keep all loci)
indivs1 <- c("Pb2Jp", "Pb2Scx", "Pb1Rd")
vcf_outgrp <- vcf_extract_samples(vcf, indivs1, f_invar = FALSE)
vcf_ingrp <- vcf_extract_samples(vcf, indivs1, keep = FALSE, f_invar = FALSE)

# extract groups of individuals from VCFArrow object (keep all loci)
groups1 <- c("GS", "BS", "WA")
vcf_outgrp <- vcf_extract_groups(vcf, groups1, keep = FALSE, f_invar = FALSE)
vcf_ingrp <- vcf_extract_groups(vcf, groups1, f_invar = FALSE)

# bind vcf_outgrp and vcf_ingrp (objects must share all variants)
vcf1 <- vcf_bind(vcf_ingrp, vcf_outgrp)

# extract groups of individuals from VCFArrow object (remove invariant loci)
groups1 <- c("GS", "BS", "WA")
vcf_outgrp <- vcf_extract_groups(vcf, groups1, keep = FALSE)
vcf_ingrp <- vcf_extract_groups(vcf, groups1)

# bind vcf_outgrp and vcf_ingrp (intersection - variable SNVs in all objects)
vcf1 <- vcf_bind_sparse(vcf_ingrp, vcf_outgrp, mode = "intersect")

# bind vcf_outgrp and vcf_ingrp (intersection - variable SNVs in all objects)
vcf1 <- vcf_bind_sparse(vcf_ingrp, vcf_outgrp, mode = "union", absent_as = "missing")

```

### Filtering with outgroup

Outgroups, which tend to have many fewer individuals than ingroups, also have more missing data than ingroups due to phylogenetic effects. Filtering the entire dataset often results in the removal of the outgroup taxa when `vcf_filter_missing()` is used during filtering to remove individuals with % missing data above some threshold. It also results in fewer SNVs being retained after filtering. The solution is to extract the outgroup taxa with `vcf_extract_samples()` or `vcf_extract_groups()`, filter the ingroup VCFArrow object, and add the outgroup taxa with `vcf_bind_sparse()` using the 'union' binding option. 

```{r} 
# extract outgroup from VCFArrow
groups1 <- "OG"
vcf_outgrp <- vcf_extract_groups(vcf, groups1)

# filter VCFArrow ingroup, then bind outgroups
vcf1 <- vcf_extract_groups(vcf, groups1, keep = FALSE) |>
  vcf_filter_indels() |>
  vcf_filter_biallelic() |>
  vcf_filter_coverage(10) |>
  vcf_filter_missingness(.2) |>
  vcf_filter_missing(.3) |>
  vcf_filter_oneSNV() |>
  vcf_bind_sparse(vcf_outgrp, mode = "union", absent_as = "missing")

# see how many variants remained
nrow(vcf1@variants)

# apply the same filter to the full dataset (ingroup + outgroup)
vcf1 <- vcf |>
  vcf_filter_indels() |>
  vcf_filter_biallelic() |>
  vcf_filter_coverage(10) |>
  vcf_filter_missingness(.2) |>
  vcf_filter_missing(.3) |>
  vcf_filter_oneSNV()

# see how many variants remained
nrow(vcf1@variants)

```


## Saving and copying VCFArrow objects

The filtering steps never actually remove any sample or SNVs from the GT matrix which is a design decision. If you want a new VCFArrow object that only has the post filtering samples and SNVs, you can use the function `vcf_copy()` which creates a new object and removes any filtered samples and SNVs. After filtering the VCFArrow object you will also want to save it as a VCF. This is accomplished using the function `write_vcf()`, which, by default will save an uncompressed VCF, but setting the 'gzip' flag to TRUE will gzip compress the VCF. With large VCFArrow objects this has a large overhead, so it is more efficient to save an uncompressed VCF and then compress it later.


```{r} 
# write the VCFArrow object as a VCF (and compress it using gzip)
write_vcf(vcf1, out_file = file.path(res_path, paste0(project, postfix, "_filtered.vcf.gz")), gzip = TRUE)

```


## Converting a VCF file to other population genetic and phylogenetic formats

All the following functions will take a VCFArrow object, and convert it other populations genetic and phylogenetic formats, exporting/writing a file of this format. Group/population information is extracted from the VCFArrow object for those formats that require it. The function `vcf2genlight()` automatically returns a genlight object and optionally can also export/write a file of this format to the working directory. Generally the `vcf2genlight()` function is called within a script using functions of the `adegenet` and `poppr` packages rather than importing the genlight object.

```{r} 
res_path <- data_path
project <- "vaillantii_"
postfix <- "discosnp_sub"
fltr <- ""

##########
# datasets for analyses with unlinked SNPs
vcf <- vcf_oneSNP

##########
# export data formats
# migrate-n https://peterbeerli.com/migrate-html5/
vcf2migrate(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '_migrate.txt')))
# arlequin http://cmpg.unibe.ch/software/arlequin35/
vcf2arlequin(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.arp')))
# structure https://web.stanford.edu/group/pritchardlab/structure.html
vcf2structure(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.str')))
# faststucture http://rajanil.github.io/fastStructure/
vcf2structure(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.fstr')), method = "F")
# sNMF http://membres-timc.imag.fr/Olivier.Francois/snmf/index.htm
vcf2snmf(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.geno')))
# Admixture https://dalexander.github.io/admixture/index.html
# takes as input binary PLINK (.bed), ordinary PLINK (.ped), or EIGENSTRAT (.geno) formatted files
# PLINK .bed https://www.cog-genomics.org/plink/1.9/formats#bed
vcf2plink_bed(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '_plink')))
# PLINK .ped https://www.cog-genomics.org/plink/1.9/formats#ped
vcf2plink_ped(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '_plink')))
# eigenstrat https://github.com/DReichLab/EIG/tree/master
vcf2eigenstrat(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '_eigenstrat')))
# genepop https://gitlab.mbb.univ-montp2.fr/francois/genepop
vcf2genepop(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.gen')))
# smartsnp https://github.com/ChristianHuber/smartsnp
vcf2smartsnp(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.smartsnp')))
# bayescan https://github.com/mfoll/BayeScan
vcf2bayescan(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.bayescan')))
# bayesass https://github.com/brannala/BA3
vcf2bayesass(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.bayesass')))
# treemix https://bitbucket.org/nygcresearch/treemix/wiki/Home
vcf2treemix(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.treemix')))
# apparent https://github.com/halelab/apparent/tree/master
vcf2apparent(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.apparent')))
# related https://github.com/timothyfrasier/related
vcf2related(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.related')))
# long tidy dataframe of genotypes
vcf2gt_long(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.csv')), format = 'csv')
# snapp https://www.beast2.org/snapp/
vcf2snapp(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '_snapp.nex')))
# nexus - only SNPs, meant for SVDq analyses https://www.asc.ohio-state.edu/kubatko.2/software/SVDquartets/
vcf2nexus(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '_sdvq.nex')))
# fasta https://www.ncbi.nlm.nih.gov/genbank/fastaformat/
vcf2fasta(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.fna')))

# genlight object https://www.rdocumentation.org/packages/adegenet/versions/2.0.0/topics/genlight-class
genlight <- vcf2genlight(vcf)
# genlight object with an optional save
genlight <- vcf2genlight(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '_genlight.rds')), save = TRUE)

##########
# datasets for analyses with linked SNPs
vcf <- vcf_multiSNP

##########
# export data formats
# fineRadStructure - expects VCF of linked SNPs
vcf2fineradstructure(vcf, out_file = file.path(res_path, paste0(project, postfix, fltr, '.finerad')))

```
